The language of generalization: Probability, vagueness, and interaction

### 2018-09-22 paper/comgen-paper.Rmd
move philosophical puzzles to model section, begin working through examples, draft a table

### Worked examples

The endorsement model defined in the previous section explicitly predicts that the variability in endorsements for generalizations in language can be attributed to two factors: (i) the referent prevalence in the category under discussion $p$ in Eq. \ref{eq:S1} (e.g., $P(x \text{ lays eggs} \mid x \text{ is a robin})$) and (ii) the background knowledge about prevalences---the prevalence prior---$P(p)$ (Eq. \ref{eq:L0}).
Background knowledge may be richly structured, reflecting intuitive theories about the underlying causes of different kinds of properties [cf., @Leslie2007].
This model posits, however, that the only impact of structured knowledge on truth judgments is in their implications for beliefs about prevalence. 
We simulate how the prevalence priors change the literal interpretation and how they and the referent prevalence impact endorsement using schematic parameter values based on intuition from a few examples in the literature on generic language.
We implemented these and all subsequent Bayesian models in the probabilistic programming language WebPPL [@dippl].
All models, analyses, data and links to experiments used in this paper can be found at \url{https://mhtess.github.io}.

Figure\ \@ref(fig:uniformPriorSimulation) shows different components of the model behavior assuming a uniform prevalence prior $P(p)$. 
The literal interpretation model (Eq. \ref{eq:L0}) computes a posterior distribution on prevalence by considering different possible thresholds $\theta$.
If the threshold were to be very high (e.g., $\theta = 0.75$), only the highest prevalence levels would be consistent with the utterance (Figure\ \@ref(fig:uniformPriorSimulation)A, right).
As the threshold decreases, more and more prevalence levels would be true  (Figure\ \@ref(fig:uniformPriorSimulation)A).
Roughly speaking, the uncertain-threshold interpretation model averages over these possibilities, resulting in a posterior distribution that favors higher prevalence levels because they are consistent with more thresholds (Figure\ \@ref(fig:uniformPriorSimulation)B).
The endorsement model inverts the interpretation model and predicts higher rates of endorsement as the speaker's belief about the prevalence (referent-prevalence) increases (Figure\ \@ref(fig:uniformPriorSimulation)C).

<!-- The literal interpretation of a generalization (Eq. \ref{eq:L0}) reflects a balance between an uncertain threshold and shared background knowledge about the feature in question. -->

<!-- TELL US WHAT THEY ARE: BETA DISTRIBUTIONS./ -->
The uncertain-threshold model's interpretations are highly sensitive to the shape of the prevalence priors (Figure\ \@ref(fig:simulations)).
Prior distributions over prevalence can be revealed by considering the prevalence of the feature in alternative categories.
Think about five of your favorite kinds of animal: What percentage of each of them *are female*? What percentage of each of them *lay eggs*?
Figure\ \@ref(fig:simulations) (row 1) show hypothetical prevalence priors for six different properties, formalized as Beta distributions or mixtures of Beta distributions.^[
  The Beta distribution is a distribution over numbers between 0 and 1. 
  In other words, it is a probability distribution over probabilities or prevalences.
]
The *lay eggs* distribution has substantial probability mass near 0 (since many animal categories do not have egg-layers) with a secondary component peaked around 0.5, because among the animals with egg-layers, only the female members of the category have the property.
Conversely, the distribution over the prevalence of *being female* is unimodal and centered at 0.5, because almost all animals have female members in that proportion.

Hearing "Ks bark" tells the literal interpreter that *almost all* Ks bark, though there is a good deal of uncertainty in the precise quantity (column 1).
"Ks lay eggs" has a strikingly different interpretation: It means "Half of Ks lay eggs" (column 4).
The interpretation model can also derive very weak interpretation, as shown in the  *carries malaria* distribution (column 6).

@@

@@

## Other philosophical puzzles

**1. Mary handles the mail from Antarctica (*yet has never had the opportunity*).**

**2. Supreme Court Justices have even social security numbers.**

**3. Elephants live in Africa and Asia.**

>>>>

### Model behavior

The endorsement model defined in the previous section predicts that the variability in endorsements for generic statements and other sorts of generalizations in language can be attributed to two factors: (i) the referent prevalence in the category under discussion $p$ in Eq. \ref{eq:S1} (e.g., $P(x \text{ lays eggs} \mid x \text{ is a robin})$) and (ii) the background knowledge about prevalences---the prevalence prior---$P(p)$ (Eq. \ref{eq:L0}).
We explore the predictions of this model by varying each of these factors in the context of a few worked examples.
These examples are taken from the linguistics literature on generics and are statements that any theory of generics should be able to correctly predict are either true, false, or somehow indeterminate. 
We implemented these and all subsequent Bayesian models in the probabilistic programming language WebPPL [@dippl].
All models, analyses, data and links to experiments used in this paper can be found at \url{https://mhtess.github.io}.

<!-- Later we will test these predictions in a quantitative way using empirical data. -->
<!-- We simulate how the prevalence priors change the literal interpretation and how they and the referent prevalence impact endorsement using schematic parameter values based on intuition from a few examples in the literature on generic language. -->
First, we examine the model behavior assuming a uniform prevalence prior $P(p)$, representing a purely abstract property for which an interpreter has no substantive prior knowledge (Figure\ \@ref(fig:uniformPriorSimulation))
The literal interpretation model (Eq. \ref{eq:L0}) computes a posterior distribution on prevalence by considering different possible thresholds $\theta$.
If the threshold were to be very high (e.g., $\theta = 0.75$), only the highest prevalence levels would be consistent with the utterance (Figure\ \@ref(fig:uniformPriorSimulation)A, right).
As the threshold decreases, more and more prevalence levels become be compatible with the threshold (Figure\ \@ref(fig:uniformPriorSimulation)A).
Roughly speaking, the uncertain-threshold interpretation model averages over these possibilities, resulting in a posterior interpretation distribution that favors higher prevalence levels because they are consistent with more thresholds (Figure\ \@ref(fig:uniformPriorSimulation)B).
The endorsement model then inverts the interpretation model and predicts higher rates of endorsement as the speaker's belief about the prevalence (referent-prevalence) increases (Figure\ \@ref(fig:uniformPriorSimulation)C).

<!-- The literal interpretation of a generalization (Eq. \ref{eq:L0}) reflects a balance between an uncertain threshold and shared background knowledge about the feature in question. -->
<!-- TELL US WHAT THEY ARE: BETA DISTRIBUTIONS./ -->

The uncertain-threshold model's interpretations are highly sensitive to background knowledge.
Background knowledge may be richly structured, reflecting intuitive theories about the underlying causes of different kinds of properties [cf., @Leslie2007].
This model posits, however, that the only impact of structured knowledge on truth judgments is in their implications for beliefs about prevalence. 
Prior distributions over prevalence can be revealed by considering the prevalence of the feature in alternative categories (Figure\ \@ref(fig:simulations)).
Think about five of your favorite kinds of animal: What percentage of each of them *are female*? What percentage of each of them *lay eggs*?
Figure\ \@ref(fig:simulations) (row 1) show hypothetical prevalence priors for six different properties, formalized as Beta distributions or mixtures of Beta distributions.^[
  The Beta distribution is a distribution over numbers between 0 and 1. 
  In other words, it is a probability distribution over probabilities or prevalences.
]
The *lay eggs* distribution has substantial probability mass near 0 (since many animal categories do not have egg-layers) with a secondary component peaked around 0.5, because among the animals with egg-layers, only the female members of the category have the property.
Conversely, the distribution over the prevalence of *being female* is unimodal and centered at 0.5, because almost all animals have female members in that proportion.

```{r workedExamples, results="asis"}
worked.examples <- data.frame(
  example = c("1. Dogs bark",
              "2. Kangaroos have spots",
              "3. Sharks don't eat people",
              "4. Robins lay eggs",
              "5. Robins are female",
              "6. Mosquitos carry malaria", 
              "7. Mary handles the mail from Antarcita",
              "8. Supreme Court Justices have even social security numbers",
              "9. Elephants live in Africa and Asia"),
  `referent prevalence` = c("95%", "5%", "85%", "50%", "50%", "5%", "0 times", "100%", "~50%/50%"),
  `intuitive truth value` = c(
    "T", "F", "F", "T", "U",
    "T", "T", "F", "T" 
  ),
  `potential issue` = c(
    "Not all dogs bark",
    "Some kangaroos could have spots",
    "Most sharks don't eat people",
    "Only female (50%) of robins lay eggs",
    "Roughly same number of robins that lay eggs are female",
    "Very few mosquitos actually carry malaria",
    "Hypothetically, there has never been any mail from Antarcita",
    "Even if 100% have even social security numbers",
    "Impossible for most to live in Africa and most to live in Asia"
  )
)

examples <- xtable::xtable(worked.examples, 
                       caption = "Example generalizations that a theory of generics should correctly predict.", 
                       align=c('p{1.5in}', 'p{1.2in}| ',' |p{1in}|','p{1in}|','p{2in} |'))

addtorow          <- list()
addtorow$pos      <- list()
addtorow$pos[[1]] <- c(0)
addtorow$command  <- c(paste("\\hline \n",
                             "\\endhead \n",
                             "\\hline \n",
                             "{\\footnotesize \\em{continued on next page}} \n",
                             "\\endfoot \n",
                             "\\endlastfoot \n",sep=""))

large <- function(x){ paste0('{\\Large ', x, '}') }
bold <- function(x){ paste0('{\\bfseries ', x, '}') }

print(examples, 
      type="latex", 
      comment = F,
       #table.placement = "h", 
      tabular.environment = "longtable",
      hline.after=-1:nrow(examples),
      size="\\fontsize{9pt}{10pt}\\selectfont", 
      include.rownames=FALSE,
      sanitize.colnames.function = bold
      #add.to.row = addtorow
      )
```

#### Worked examples

We are now ready to examine the endorsement model's predictions for different examples.
These examples have historically proved challenging for prevalence-based approaches. 
We describe how the uncertain-threshold model treats each in turn.

**1. Dogs bark.**

Often the first observation with generics is that they appear like *universal* quantifiers (e.g., "All") but permit exceptions. 
That is, not all dogs bark, but still *Dogs bark* is true.
The uncertain threshold account immediately accomodates this example because there is no single, fixed threshold beyond which the generalization becomes true.
Rather, listeners have uncertainty about the threshold *a priori* which leads to graded interpretations (e.g., *Dogs bark* meaning *almost all* dogs bark; Figure\ \@ref(fig:simulations); column 1, row 2).
The endorsement model then predicts this is a rather good generic sentence (Figure\ \@ref(fig:simulations), bottom row) even though not all dogs bark (shown with arrow in Figure\ \@ref(fig:simulations) row 2).

**2. Kangaroos have spots.**

A second observation regarding generics is that they cannot be so lenient so to always convey *existential* quantification (e.g., "Some").
Indeed, it's intuitively plausible that some kangaroos do have spots but a rational language user might feel the generalization "Kangaroos have spots" is too strong to assert.
The model exhibits this same restraint, predicting a very low endorsement probability for this generic statement (Figure\ \@ref(fig:simulations); column 2).
The reason is that, given the interpretation model's background knowledge about the property, the statement "Kangaroos have spots" is like "Dogs bark": it means *almost all* kangaroos have spots. 
The endorsement model, which believes that very few kangaroos have spots, then would rather not endorse the statement because it would mislead the listener.

**3. Sharks don't eat people.**

A similar logic can be applied to this example, which is the converse of the affirmative generalization "Sharks eat people", initially introduced by @Leslie2007. 
Here, however, the prevalence of the feature for the referent category---the number of sharks that don't eat people---is very high: Almost all sharks don't eat people.
Yet, intuitively, the generalization is again misleading. 
Because for almost all animal categories, almost members of those categories don't eat people (i.e., the prevalence of the feature is almost 100% for all categories because very few things eat people), interpreting the statement "Sharks don't eat people" would lead one to believe that no sharks eat people, which is too strong.
It is interesting to note that the model is less certain of its decision, predicting an endorsement probability around 40%.
We will test this quantitative prediction in one of our experiments.

**4. and 5. Robins lay eggs vs. Robins are female**

**6. Mosquitos carry malaria.**

Hearing "Ks bark" tells the literal interpreter that *almost all* Ks bark, though there is a good deal of uncertainty in the precise quantity.
"Ks lay eggs" has a strikingly different interpretation: It means "Half of Ks lay eggs" (column 4).
The interpretation model can also derive very weak interpretation, as shown in the  *carries malaria* distribution (column 6).

@@

#### Other philosophical puzzles

**7. Mary handles the mail from Antarctica (*yet has never had the opportunity*).**

**8. Supreme Court Justices have even social security numbers.**

**9. Elephants live in Africa and Asia.**

@@
